---
title: 'Accelerated Lecture 4: If statements and conditionals'
author: "Harris Coding Camp -- Standard Track"
date: "Summer 2022"
output:
  beamer_presentation:
    fig_width: 6
    fig_height: 4
    includes:
      in_header: ../resources/header_page_n.tex
---
```{r, setup=TRUE, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readr)
options(dplyr.summarise.inform = FALSE) # Suppress summarize info in dplyr >= 1.0
```

## Review: Subsetting data

```{r, eval = FALSE}
# base R
data[row_condition, c("columns", "we", "want"]

# tidyverse
data %>%
  filter(row_condition) %>%
  select(columns, we, want)
```


\footnotesize
*Remark*: You may also see base R's `subset()`


## Review: Sorting data

```{r, eval = FALSE}
# base R
data[order(data$col, -data$col2),]

# tidyverse
arrange(data, col, desc(col2))
```


## Review: Summarizing data

```{r, eval = FALSE}
# base R
# results in a vector of length 1
mean(data$col)

# tidyverse 
# results in a tibble with 1 row
summarize(data, mean = mean(col))

```

\footnotesize
- We let `mean` stand in for any function that reduces the data to a single value (per group)
- *Remark* base R code can get more sophisticated

## Review: Creating new data


```{r, eval = FALSE}
# base R
data$new_column <- something 

# tidyverse
data <- data %>% mutate(new_column = something) 
```


Same functionality to change old data:

```{r, eval = FALSE}
# base R
data$old_column <- something 

# tidyverse
data <- data %>% mutate(old_column = something) 
```

\footnotesize
- `something` is length nrow(data) or 1

## How would we make a column dependent on other data?
\footnotesize
```{r, echo = FALSE}
set.seed(12)
d <- tibble(x = 1:4,
       y = round(rnorm(4), 1))

d
```

Add column dependent on `y`
```{r, echo = FALSE}
d %>% mutate(set_neg_y_to_0 = -y * (y < 0) + y)
```




# Call in `if` and `ifelse`

## Conditional statements

We often want to our code to do something depending on the context. We start with `if` statements.

```{r eval=FALSE}
if (condition is true) {
  do this
} else {                      
  do this other thing
}
```

We will cover:

 - introduce `if` and `else` statements
 - introduce vectorized `ifelse` and `case_when()` statements 
    

## `if` statements

The general syntax of an `if` statement is as follows:
```{r eval=FALSE}
if (condition is TRUE) {
  do this
}
```

For example:
```{r}
x <- 100

if (x > 0) {
  print("x is positive")
}

```

## `if/else` statements

Slightly more interesting, the syntax of an `if/else` statement is as follows:

```{r eval=FALSE}
if (condition is TRUE) {
  do this
} else {                      
  do this other thing
}
```

## `if/else` statements, example

```{r}
x <- -5
if (x > 0) {
  print("Non-negative number")
} else {
  print("Negative number")
}
```

## `if`, `else if` and `else` statements

If we have more than 2 conditions, use `if`, `else if` and `else`:
```{r eval=FALSE}
if (condition is TRUE) {
  do this
} else if (second condition is TRUE) {                      
  do this other thing
} else if (third condition is TRUE) {
  do this third thing
} else {
  do a default behavior
}
```

Note: a default behavior with `else` is not necessary.

## `if`, `else if` and `else` statements, example

\footnotesize
```{r}
x <- sample(1:100, 1)
x
y <- sample(1:100, 1)
y
```

\footnotesize
```{r}
if (x > y) {
  print("x is greater")
} else if (x < y) {
  print("y is greater")
} else {                
  print("x and y are equal")
}
```

## `if`, `else if` and `else` can take a compound condition

```{r}
x <- sample(1:100, 1)
x
y <- sample(1:100, 1)
y
z <- sample(1:100, 1)
z
```

## `if`, `else if` and `else` can take a compound condition

```{r}
if (x >= y & x >= z) {
  print("x is the greatest")
} else if (y >= z) {
  print("y is the greatest")
} else {                
  print("z is the greatest")
}
```

## Try it yourself
\footnotesize
Let's develop a small dice game. 

1. Fill in the ... so the code says "You win" if the dice add up to 7 and "You lose" otherwise.

```{r, eval = FALSE}
dice <- sample(c(1:6), 2)
    
if (...) {
  print("You win")
} else {
  print("You lose")
}
```

2. Add an `else if()` block to the code above that says "try again" if the dice add up to 6 or 8.


## Try it yourself
\footnotesize
2. Add an `else if()` block to the code above that says "Try again" if the dice add up to 6 or 8.

```{r, eval = FALSE}
dice <- sample(c(1:6), 2)
    
if (...) {
  print("You win")
} else if (...) {
  print("Try again")
} else {
  print("You lose")
}

```

```{r, eval = FALSE, echo=FALSE}
# answer
dice <- sample(c(1:6), 2)

if (sum(dice) == 7) {
  print("You win")
} else if(sum(dice) %in% c(6,8)) {
  print ("Try again")
} else {
  print("You lose")
}

```


## Some common uses of `if`

Sharing code among various people. 

- `Sys.getenv("USER")` returns the name of the USER fr

```{r, eval = FALSE}

if (Sys.getenv("USER") == "arianisfeld") {
  setwd("~/repo/dir")
} else if (Sys.getenv("USER") == "yunjoo") {
  setwd("C://repo/dir")
} else {
  print(paste0("WARNING: Unknown user. 
               Working directory is ", getwd()))
}

```

## Some common uses of `if`

In a function, you might want to adjust to different inputs.

*example:* How can we code up the absolute value function?

$$|x| = \begin{cases}x, & x >= 0\\ -x,  & x <0\end{cases}$$


## Some common uses of `if`

In a function, you might want to adjust to different inputs.

```{r}
absolute_value <- function(x) {
  # x: a numeric of length 1
  
  if (x < 0) {
    
    x <- x * -1
  
  }
  
  return(x)
}
```

## `if()` the condition must return TRUE or FALSE

`if()` is not vectorized!

```{r, eval = FALSE}
x <- c(1, -4)

if (x > 0) {
   x
} else {
  -x
}
```

`Error in if (x > 0) { : the condition has length > 1`

- This error is as of R 4.2.0


## `if()` the condition must return TRUE or FALSE


`if()` does not handle `NA`s

```{r, eval = FALSE}
x <- NA

if (x > 0) {
  x
} else {
  -x
}
```

`Error in if (x > 0) { : missing value where TRUE/FALSE needed`


## `if()` the condition must return TRUE or FALSE
\footnotesize
You may need to write code to handle edge cases.

```{r, eval = TRUE}
x <- NA

if (all(length(x) == 1 & !is.na(x) & x > 0)) {
  x
} else if (length(x) == 1) {
  -x
} 


x <- c(123, 1)
out <- if (all(length(x) == 1 & !is.na(x) & x > 0)) {
  x
} else if (length(x) == 1) {
  -x
} 
out
```

## Detour: `NULL` vs `NA`
\footnotesize
`NULL` stands in for an *object* that is undefined.

```{r}
length(NULL)
NULL > 1
```  

~~~

`NA` stands in for a *value* that is undefined.

```{r}
length(NA)
NA > 1
```  



## `if()` the condition must return TRUE or FALSE
\footnotesize
Good idea to make sure it still works for valid input!

```{r}
x <- 1309
if (all(length(x) == 1 & !is.na(x) & x > 0)) {
  x
} else if (length(x) == 1) {
  -x
} 


x <- -1 * pi
if (all(length(x) == 1 & !is.na(x) & x > 0)) {
  x
} else if (length(x) == 1) {
  -x
} 

```




## Vectorized `ifelse` statements
\footnotesize
In R, the `ifelse()` function is a shorthand vectorized alternative to the standard `if...else` statement.

Syntax: `ifelse(test, yes, no)` 

```{r}
x <- 5
y <- 50
ifelse(x > y, "x is greater", "x is not greater")
```

```{r}
a <- 60
b <- 6
ifelse(a > b, "a is greater", "a is not greater")
```

## What will the following statements return? 

```{r eval=FALSE}
ifelse(TRUE, 1, 2)
ifelse(FALSE, 1, 2)
```

## What will the following statements return?

```{r eval=TRUE}
ifelse(TRUE, 1, 2)
ifelse(FALSE, 1, 2)
```

## What will the following statements return? 

```{r eval=FALSE}
ifelse(c(TRUE, FALSE, FALSE, TRUE), 1, 2)

ifelse(1:4 > 3, 1, 2)

```

## What will the following statements return? 

```{r eval=TRUE}
ifelse(c(TRUE, FALSE, FALSE, TRUE), 1, 2)

ifelse(1:4 > 3, 1, 2)

```

## `ifelse` handles NAs and missing data

What's going on in this `ifelse()` statement?

```{r echo=TRUE}
ifelse(NA, 1, 2)
ifelse(NULL, 1, 2)
```

- `NA`s are contagious.

## `ifelse` statements in dataframes, base R

`ifelse` statements work well in dataframes when we need to create a new column.

- Let's add a column to the `txhousing` based on a conditional.
\footnotesize
```{r}
txhousing$in_january <- 
  ifelse(txhousing$month == 1, TRUE, FALSE) 
```

## `ifelse` statements in dataframes for multiple categories

If we have a lot of categories, use **nested** `ifelse` statement. \newline Say we want to create a new variable called `median_ref` which value can be High, Medium or Low:

  - If `median` is at least 70k: High 
  - If `median` is between 60k and 70k: Medium 
  - If `median` is lower than 60k: Low
  
\footnotesize
```{r}
txhousing$median_ref <- 
  ifelse(txhousing$median >= 70000, 'High', 
         ifelse(txhousing$median < 70000 & 
                txhousing$median >= 60000, 
                'Medium', 'Low')) 
```

## `ifelse` statements in dataframes, tidyverse
\footnotesize

Use `ifelse` statements in `mutate()` function.

- Let's add a column called `in_january` to the `txhousing` based on a condition.

```{r}
txhousing %>% 
  mutate(in_january = ifelse(month == 1, TRUE, FALSE)) %>% 
  select(city, year, month, sales, in_january)
```


## `ifelse` statements in dataframes, tidyverse 
\footnotesize
As before, we can handle nested statements with `ifelse()`

```{r}
txhousing %>% 
  select(city, year, month, median) %>%
  mutate(housing_market = 
      ifelse(median < 100000, "first quartile",
      ifelse(100000 <= median & median < 123800, "second quartile",
      ifelse(123800 <= median & median < 150000, "third quartile",
      ifelse(150000 <= median & median < 350000, "fourth quartile",
             NA))))
           ) %>%
  head(3)
```




## `case_when` statements in dataframes, tidyverse 
\footnotesize

There's a cleaner way to handle multiple cases.

- Instead of nesting `ifelse` statements we can use `case_when()` 

```{r eval=FALSE}
# add a column called `housing_market` to the `txhousing` 
txhousing %>% 
  select(city, year, month, median) %>%
  mutate(housing_market = 
         case_when(
           median < 100000 ~ "first quartile",
           100000 <= median & median < 123800 ~ "second quartile",
           123800 <= median & median < 150000 ~ "third quartile",
           150000 <= median & median < 350000 ~ "fourth quartile"
         ))  %>%
  head(3)
```


## `case_when` statements are a bit "surly"
\footnotesize
`case_when` will not do type coercion. 

```{r, eval = FALSE}
txhousing %>% 
  mutate(housing_market = 
        case_when(
           median < 100000 ~ 1,
           100000 <= median & median < 123800 ~ "second quartile",
           123800 <= median & median < 150000 ~ "third quartile",
           150000 <= median & median < 350000 ~ "fourth quartile"
         )) %>% 
  select(city, median, housing_market)

Error: must be a double vector, not a character vector
Run `rlang::last_error()` to see where the error occurred.
```

Here we try to include *both* doubles and characters in the `housing_market` column, but atomic vectors can only have one type!

  - Rather than coerce and provide a warning, the developers decided to make this an error
  - If using `NA` as an output, you have to specify `NA` types e.g. `NA_integer_`, `NA_character_`

## Try it yourself
\footnotesize
We will use `midwest` here, which is a dataset built into `tidyverse`.

1. Create a new variable called `poverty_designation` that is "High Poverty" if `percbelowpoverty` is above 10 and is "Low Poverty" otherwise.

```{r, eval=FALSE, echo = FALSE}
# answer
midwest$poverty_designation <- ifelse(midwest$percbelowpoverty > 10, "High Poverty", "Low Poverty") 
```

2. Create a new variable called `ohio` that is "Ohio Counties" for observations from Ohio and "Other Midwestern Counties" for the rest of the observations.

```{r, eval=FALSE, echo = FALSE}
# answer
midwest$ohio <- ifelse(midwest$state == "OH", "Ohio Counties", "Other Midwestern Counties") 
```

3. Create a new variable called `populous_counties` that is `TRUE` for the observations from the counties listed in `big_counties` and `FALSE` otherwise. Hint: Use the `%in%` operator.\footnotesize
```{r, eval = FALSE}
big_counties <- c("COOK", "WAYNE", "CUYAHOGA", "OAKLAND", "FRANKLIN")
```

```{r, eval=FALSE, echo = FALSE}
# answer
big_counties <- c("COOK", "WAYNE", "CUYAHOGA", "OAKLAND", "FRANKLIN")
midwest$populous_counties <- midwest$county %in% big_counties
```

4. Create a new variable called `pop_index` that is "High" for the observations with `poptotal` greater than 100000, is "Medium" for the observations with `poptotal` between 30000 and 100000, and "Low" otherwise.

```{r, eval=FALSE, echo = FALSE}
# answer
midwest$pop_index <- 
  ifelse(midwest$poptotal > 100000, "High", 
    ifelse(midwest$poptotal >= 30000 & midwest$poptotal <= 100000, "Medium", 
      "Low")
```


## Recap

Today we learned how to:

- use control flow with `if` and `ifelse` statements
- use `ifelse()` and `case_when()` statements in conjunction with `mutate()` or `$<-` to create columns based on conditional statements

## Next up

Labs:

- Today: Practice with `ifelse`
- Tomorrow: Coding style, review and catch-up.

**I can use `ifelse` to create columns conditional on data**

and

**I'm gaining confidence doing basic data manipulation**

Lecture:

- Making data visualizations

