---
title: 'Accelerated Lecture 7: Writing Functions'
author: "Harris Coding Camp"
date: "Summer 2022"
output:
  beamer_presentation:
    includes:
      in_header: ../resources/header_page_n.tex
---
```{r, setup=TRUE, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
#options(dplyr.summarise.inform = FALSE)
#knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE, fig.align="center")
```

## Functions

```{r}
# example of a function
circle_area <- function(r) {
   pi * r ^ 2
}
```

 - What are functions and why do we want to use them?
 - How do we write functions in practice?
 - What are some solutions to avoid frustrating code?
 
## Motivation

- Grolemund and Wickham chapter 19:
    - "You should consider writing a function whenever you’ve copied and pasted a block of code more than twice (i.e. you now have three copies of the same code)."

- Darin Christenson refers to the programming mantra __DRY__
    - "Do not Repeat Yourself (DRY)"
    - "Functions enable you to perform multiple tasks (that are similar to one another) without copying the same code over and over"

## Instead of repeating code...
\footnotesize
```{r, echo = FALSE}
data <-
tibble(a = rnorm(100),
       b = rnorm(100),
       c = rnorm(100),
       d = rnorm(100))
```

```{r issue}
data %>%
  mutate(a = (a - min(a)) / (max(a) - min(a)),
         b = (b - min(b)) / (max(b) - min(b)),
         c = (c - min(c)) / (max(c) - min(c)),
         d = (d - min(d)) / (max(d) - min(d)))
```

## Write a function
\footnotesize
```{r}
rescale_01 <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

data %>%
  mutate(a = rescale_01(a),
         b = rescale_01(b),
         c = rescale_01(c),
         d = rescale_01(d))
```

## Function anatomy
\footnotesize
The anatomy of a function is as follows:

```{r eval=FALSE}
function_name <- function(argument_1, argument_2) {
  do_this(argument_1, argument_2)
}
```

Three components of a function:

1. __function name__
    - specify function name before the assignment operator `<-`
2. __function arguments__ (sometimes called "inputs" or "arguments")
    - Inputs that the function takes
        - can be vectors, data frames, strings, etc.
    - In above hypothetical code, the function took two inputs `argument_1`,`argument_2`
    - In "function call," you specify values to assign to these function arguments
3. __function body__
    - What the function does to the inputs

## Function anatomy: example

+ **arguments**: `x`
+ **body**: `(x - min(x)) / (max(x) - min(x))`
+ assign to **name**: `rescale_01`

```{r}
rescale_01 <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
```

The last line of the code will be the value returned by the function!

- We do not explicitly call `return()`


## Writing a function: printing output

You start writing code to say Hello to all of your friends.

- You notice it's getting repetitive. ... time for a function!

```{r}
print("Hello Jasmin!")
print("Hello Joan!")
print("Hello Andrew!")
# and so on...
```

## Writing a function: parameterize the code

Start with the **body**.

Q: What part of the code is changing? Or what aspects of the code do you want to change?

 - Make this an **argument**
 
## Writing a function: parameterize the code

Start with the **body**.

(Re)write the code to accommodate the parameterization:

```{r}
# print("Hello Jasmin!") becomes ...

name <- "Jasmin" 

print(paste0("Hello ", name, "!"))
```

Check several potential inputs to avoid future headaches

## Writing a function: add the structure

Now let's add the **structure** to formally define the new function:

```{r, eval = FALSE}
# name <- "Jasmin"
# print(paste0("Hello ", name, "!"))

function(name) {
  print(paste0("Hello ", name, "!"))
}
```

+ **arguments**: `name`
+ **body**: `print(paste0("Hello ", name, "!"))`
+ assign to **name**: not yet...

## Writing a function: assign to a name

Try to use **names** that actively tell the user what the code does 

  - We recommend `verb_thing()`
  
     - **good**: `calc_size()` or `compare_prices()`  
     - **bad**: `prices()`, `calc()`, or `fun1()`.

```{r, eval = TRUE}
# name <- "Jasmin"
# print(paste0("Hello ", name, "!"))

say_hello_to <- function(name) {
  print(paste0("Hello ", name, "!"))
}
```

+ **arguments**: `name`
+ **body**: `print(paste0("Hello ", name, "!"))`
+ assign to **name**: `say_hello_to`

## First example: printing output

Test out different inputs!

```{r}
say_hello_to("Jasmin")
say_hello_to("Joan")
say_hello_to("Andrew")

# Cool this function is vectorized!
say_hello_to(c("Jasmin", "Joan", "Andrew"))
```

## Second example: calculating the mean of a sample

Your stats prof asks you to simulate the law of large numbers, by calculating the mean of i.i.d. samples with increasing sample sizes.^[You will learn about this in Stats I]



## Recall `rnorm(n)` it generates a random sample of size n

\footnotesize
We want to: calculate the mean of i.i.d. samples with increasing sample sizes.

```{r}
rnorm(1)
rnorm(5)
rnorm(30)
```

## Too much copy paste
\footnotesize
We want to: calculate the *mean* of i.i.d. samples with increasing sample sizes.

```{r}
mean(rnorm(1))
mean(rnorm(5))
mean(rnorm(30))
# et cetera
```

## Second example: calculating the mean of a sample

The sample size is changing, so it becomes the **argument**: 

```{r}
calc_sample_mean <- function(sample_size) {
  mean(rnorm(sample_size))
}
```

  - I call the sample size `sample_size`. 
    - `n` would also be appropriate.
  - The **body** is otherwise identical to the code you already wrote.


## Function at work
\footnotesize
We want to: calculate the *mean* of i.i.d. samples with increasing sample sizes.

- This seems like the same amount of copy paste as before. (We'll improve this soon.)
- We're still better off. If we change our mind about something or find a bug, we only have to fix it once!
  
```{r}
calc_sample_mean(1)
calc_sample_mean(5)
calc_sample_mean(30)
```


## Commenting functions with clear names

For added clarity, you can unnest your code and assign the intermediate results to meaningful names:

```{r}
calc_sample_mean <- function(sample_size) {
  
  our_sample <- rnorm(sample_size)
  sample_mean <- mean(our_sample) # <- probably overkill
  
  sample_mean
  
  }
```


## Using `return()`

The last line of code run is returned by default.

- Occassionally you'll want to specify what to return

```{r}
calc_sample_mean <- function(sample_size) {
  return(mean(rnorm(sample_size)))
}
```

`return()` explicitly tells R what the function will return. 

  - Style guide says only use `return()` to break out of a function early.

## Capturing assigned output

If the last line is an assignment there is no *visible* output

- Avoid this.
\footnotesize

```{r}
calc_sample_mean <- function(sample_size) {
  sample_mean <- mean(rnorm(sample_size)) 
}

# looks like nothing happened
calc_sample_mean(1)
# but we can capture the output with an assignment
x <- calc_sample_mean(1e6)
x
```


## One-liners and anonymous functions

If the function can be fit on one line, you can write it without the curly brackets:

```{r}
calc_sample_mean <- function(n) mean(rnorm(n))
```

Some settings call for *anonymous functions*, where the function has no name.

```{r eval=FALSE}
function(n) mean(rnorm(n))
```

## Always test your code 
\footnotesize
Try to foresee the kind of input you expect to use.

```{r}
calc_sample_mean(0)
calc_sample_mean(1e6)
```

We see below that this function is not vectorized. We hoped to get 3 sample means out but only got 1

```{r}
# read ?rnorm to understand how rnorm interprets vector input
calc_sample_mean(c(1, 3, 30))
```

## How to deal with unvectorized functions
\footnotesize
If we don't want to change our function, but we want to use on vectors, then we have a couple options

 - Here we are going to use the function `rowwise()`:

```{r}
# create a tibble to test our function
sample_tibble <- tibble(sample_sizes = c(1, 3, 10, 30))

# rowwise groups the data by row
# then our function is applied to each "group"
sample_tibble %>%
  rowwise() %>%
  mutate(sample_means = calc_sample_mean(sample_sizes))
```

## Complicating the matter.


The Stats professor now calls for different paramaterizations of the normal distribution. 

- They want you to re-run your analysis with normals with different means and variances!


## Adding additional arguments
\footnotesize
If we want to be able to adjust the details of how our function runs, we can add arguments

 - typically, we put "data" arguments first
 - and then "detail" arguments after

```{r}
calc_sample_mean <- function(sample_size, 
                             our_mean, 
                             our_sd) {

  sample <- rnorm(sample_size, 
                  mean = our_mean, 
                  sd = our_sd)
  
  mean(sample)
}
```

## Setting defaults
\footnotesize
If there's a "natural" default, we can set default values for "detail" arguments


```{r}
calc_sample_mean <- function(sample_size, 
                             our_mean = 0, our_sd = 1) {
  
  sample <- rnorm(sample_size, mean = our_mean, sd = our_sd)
  
  mean(sample)
}
```

```{r}
# uses the defaults
calc_sample_mean(sample_size = 10)
```

## Setting defaults

```{r}
# we can change one or two defaults. 
# You can refer by name, or use position
calc_sample_mean(10, our_sd = 2)
calc_sample_mean(10, our_mean = 6)
calc_sample_mean(10, 6, 2)
```

## Setting defaults

This won't work though:

  - the most important argument is missing!
  
```{r eval=FALSE}
calc_sample_mean(our_mean = 5)
```

```
Error in rnorm(sample_size, mean = our_mean, sd = our_sd) : 
  argument "sample_size" is missing, with no default
```
## Another complication

Now your curious about extremes. What happens to the max as we increase our sample size?

## Functions as arguments to other functions
\footnotesize
**Before**

```{r}
calc_sample_mean <- 
  function(sample_size, our_mean = 0, our_sd = 1) {
  
  sample <- rnorm(sample_size,
                  mean = our_mean, 
                  sd = our_sd)
  mean(sample)
}
```

## Functions as arguments to other functions
\footnotesize
**After**

```{r}
summarize_sample <- function(sample_size, 
                             our_mean = 0, 
                             our_sd = 1, 
                             summary_func = mean) {
  
  sample <- rnorm(sample_size, 
                  mean = our_mean, 
                  sd = our_sd)
  
  summary_func(sample)
}
```

 Use a new descriptive name: `summarize_sample()` 
 
 - `calc_sample_mean()` would be a misleading name
 


## Functions in functions, in action
\footnotesize
```{r}
# mean(rnorm(10, mean = 0, sd = 1))
summarize_sample(sample_size = 10) # mean is default!

# max(rnorm(10, mean = 0, sd = 1))
summarize_sample(sample_size = 10, summary_func = max)

# min(rnorm(10, mean = 0, sd = 1))
summarize_sample(sample_size = 10, summary_func = min)
```

 


##  You try
1. What will happen if we execute the code below? Explain?
```{r eval=FALSE}
new.function <- function(a, b) {
   print(a)
   print(b)
}

new.function(6)
```

##  You try

2. Write a function that takes a vector and replaces negative values with `NA`. (Some data use -99 or similar numbers to represent missing-ness.)

```{r, echo = FALSE}
replace_neg <- function(x) {
  ifelse(x < 0, NA, x)
}
```

```{r}
replace_neg(c(-19, 1, 2, 1))
```

3. Write examples where you use `replace_neg()` with columns in a tibble.

```{r, include=FALSE}
tibble(x = c(-3, 1, 33, 4)) %>%
  mutate(x = replace_neg(x))
```

4. Does your function work on non-numeric inputs?

## Aside: Wait ... what?

```{r}
# +a < 0  => TRUE
# -b < 0  => TRUE
# c < 0   => FALSE
# d < 0   => FALSE

replace_neg(c("+a", "-b", "c", "d"))
```


Conditional operators work with characters! 
  
  - They test alphabetical order
  - And treat special chars and punctuation as negative!
  - So, in a real sense, you are greater than `R`

```{r}
"you" > "R"
```


## Review: Conditional execution

__`if`__ statements allow you to conditionally execute certain blocks of code depending on whether a condition is satisfied

```{r, eval=FALSE}
if (condition) {
  # code executed when condition is TRUE
} else {
  # code executed when condition is FALSE
}
```



## Suppose you don't like this behavior

We can test whether the input is numeric.

```{r}
replace_neg <- function(x) {
  if (is.numeric(x)) {
    return(ifelse(x < 0, NA, x))  
  }
}

out <- replace_neg(c("-a", "b"))
out
replace_neg(c(-3, 4))
```

## Test the function in the context you are interested in!

```{r}
# oops ... we probably don't want to lose char!
tibble(char = c("-a", "+b", "c"), 
       num = c(1, 2, -3 )) %>%
  mutate(char = replace_neg(char))
```


## Fixing the bug

```{r}
replace_neg <- function(x) {
  if (!is.numeric(x)) {
      return(x)
  }
  ifelse(x < 0, NA, x)
}
```


```{r}
# better!
tibble(char = c("-a", "+b", "c"), 
       num  = c(1, 2, -3)) %>%
  mutate(char = replace_neg(char),
         num  = replace_neg(num),)
```

## Helping a user out.
\footnotesize
We may want to warn the user that they did something funny.

```{r}
replace_neg <- function(x) {
  if (!is.numeric(x)) {
    print("Non-numeric input to replace_neg returning x")
    return(x)  
  }
  
  ifelse(x < 0, NA, x)
}

replace_neg(c("-a"))
```

- NB: `message()` makes messages and `stop()` throws errors.

## Using functions for data visualization

You're exploring data and want to make this plot for a variety of city pairings.

```{r, echo = FALSE, message = FALSE, warning=FALSE}
txhousing %>%
  filter(city %in% c("Houston", "Galveston")) %>%
  ggplot(aes(x = as_factor(year), 
             y = sales, 
             color = city)) +
  geom_boxplot(show.legend = FALSE) +
  labs(color = NULL, y = "Monthly sales", x = "Year") +
  theme_minimal() +
  facet_grid(city~., scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90))
```

## Using functions for data visualization

Suppose we want another set of plots with Austin and San Antonio. 

- Copy-paste adjust and then ... you decide you want to tinker with the plot more! Now you have to do it twice.

```{r, eval = FALSE, message = FALSE}
txhousing %>%
  filter(city == "Houston" | city == "Galveston")) %>%
  ggplot(aes(x = as_factor(year),
             y = sales, 
             color = city)) +
  geom_boxplot(show.legend = FALSE) +
  labs(color = NULL,  y = "Monthly sales", x = "Year") +
  theme_minimal() +
  facet_grid(city~., scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90))
```
## How would you parameterize this to take arbitrary cities?

```{r, eval = FALSE}
# zooming in 
txhousing %>%
  filter(city == "Houston" | city == "Galveston") ...
```



## A function for our plot

```{r}
sales_box_plot <- function(cities) {
  txhousing %>%
    filter(city %in% cities) %>%
    ggplot(aes(x = as_factor(year), 
               y = sales, 
               color = city)) +
    geom_boxplot(show.legend = FALSE) +
    labs(color = NULL,  y = "Monthly sales", x = "Year") +
    theme_minimal() +
    facet_grid(city~., scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90))
}
```

## Voila!
```{r}
sales_box_plot(c("Austin", "San Antonio"))
```
## Suppose you want `y` too ...

```{r, eval = FALSE}
housing_box_plot <- function(cities, y, ylab) {
  txhousing %>%
    filter(city %in% cities) %>%
    ggplot(aes(x = as_factor(year), 
               y = y, 
               color = city)) +
    geom_boxplot(show.legend = FALSE) +
    labs(color = NULL,  y = ylab, x = "Year") +
    theme_minimal() +
    facet_grid(city~., scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90))
}

housing_box_plot(c("Austin"), sales, "Monthly sales")
```

`Error in FUN(X[[i]], ...) : object 'sales' not found`

## Suppose you want `y` too ...

`{{...}}` embrace arguments that refer to columns in your data set.

- This is "sugar" that tells R to look within the data for the variable.

```{r, eval = TRUE}
housing_box_plot <- function(cities, y, ylab) {
  txhousing %>%
    filter(city %in% cities) %>%
    ggplot(aes(x = as_factor(year), 
               y = {{ y }}, 
               color = city)) +
    geom_boxplot(show.legend = FALSE) +
    labs(color = NULL,  y = ylab, x = NULL) +
    theme_minimal() +
    facet_grid(city~., scales = "free_y") +
    theme(axis.text.x = element_text(angle = 90))
}
```
## Voila
```{r, fig.height=2}
housing_box_plot(c("Austin"), sales, "Monthly sales")
housing_box_plot(c("Austin"), volume, "Monthly volume")
```

## Complicating `calc_sample_mean` even further.

Now you're getting really curious and want to see if these ideas hold with different distributions!

Recall:
```{r}
calc_sample_mean
```



## One approach -- make new functions for each distribution
\footnotesize
```{r}
calc_sample_mean_t <- function(sample_size, our_df) {
  sample <- rt(sample_size, our_df)
  mean(sample)
}

calc_sample_mean_chisq <- function(sample_size, our_df) {
  sample <- rchisq(sample_size, our_df)
  mean(sample)
}

# Fun fact: 2^31 -1 is the largest seed in R
set.seed(2147483647) 
calc_sample_mean_t(10,  our_df = 5)
calc_sample_mean_chisq(10, our_df = 5)
```

## A sophisticated approach - parameterize the distribution!
\footnotesize
- The complication here is each distribution has it's own parameters. `df`, `mean` etc.
- `...` takes arbitrary arguments which you can pass to another function
- Warning: `...` (dot-dot-dot) is a bit challenging to use

```{r}
calc_sample_mean <- function(sample_size, fn = rnorm, ...) {
  
  sample <- fn(sample_size, ...)
  
  mean(sample)
}


set.seed(2147483647) 
calc_sample_mean(10, rt, df = 5)
calc_sample_mean(10, rchisq, df = 5)
```

## More examples
\footnotesize
`...` takes arbitrary named arguments which you can pass to another function

```{r}
# function(sample_size, fn = rnorm, ...)
# sample <- rnorm(10)
calc_sample_mean(10)
# sample <- rf(4, df1 = 2, df2 = 3)
calc_sample_mean(4, rf, df1 = 2, df2 = 3)
# sample <- rbeta(9, shape1 = .3, shape2 = 5)
calc_sample_mean(9, rbeta, shape1 = .3, shape2 = 5)
```

## in context

\footnotesize
```{r}
tibble(x = c(1, 10, 100, 1000, 1e5)) %>%
  rowwise() %>%
  mutate(normal = calc_sample_mean(x, mean = 4, sd = 6),
         uniform = calc_sample_mean(x, runif, min = 2, max = 6),
         poisson = calc_sample_mean(x, rpois, lambda = 4))
```


## Recap

- Write functions when you are using a set of operations repeatedly
- Functions consist of arguments and a body and are usually assigned to a name
- Functions are for humans
    - pick names for the function and arguments that are clear and consistent
- Debug your code as much as you can as you write it.
    - if you want to use your code with `mutate()`, test the code with vectors
- Introduced a few sophisticated ways to work with function arguments!
    - `{{col_name}}` to refer to column names in dplyr context 
    - `...` to pass arbitrary arguments to functions.

**For more:** See [Chapter 19](https://r4ds.had.co.nz/functions.html) in R for Data Science

## Next steps:

*Lab:*

Today: Writing functions (challenging lab!)

**I can encapsulate code into functions, and debug and apply them!**

*Lecture:*

Tomorrow: Loops and iteration.


# Additional material



## Probability distributions
\footnotesize
R has built-in functions for working with distributions.

| | example | what it does? |
| -- | -- | -- | 
| r | `rnorm(n)` | generates a random sample of size n | 
| p | `pnorm(q)` | returns CDF value at q | 
| q | `qnorm(p)` | returns inverse CDF (the quantile) for a given probability |
| d | `dnorm(x)` | returns pdf value at x | 

Probability distributions you are familiar with are likely built-in to R.

For example, the binomial distribution has `dbinom()`, `pbinom()`, `qbinom()`, `rbinom()`. The t distribution has `dt()`, `pt()`, `qt()`, `rt()`, etc.


Read [this tutorial](https://thomasleeper.com/Rcourse/Tutorials/distributions.html) for more examples.

## We should be familar with `r` functions
\footnotesize
- `rnorm()`: random sampling

```{r}
rnorm(1)
rnorm(5)
rnorm(30)
```

## What are `p` and `q`?
\footnotesize
**`pnorm`** returns the probability we observe a value less than or equal to some value `q`.

```{r}
pnorm(1.96)
pnorm(0)
```

**`qnorm`** returns the inverse of `pnorm`. Plug in the probability and get the cutoff.

```{r}
qnorm(.975)
qnorm(.5)
```

This might be easier understood with pictures!

## What are `p` and `q`?

```{r, echo = FALSE, cache = TRUE}

sample <- rnorm(1e5)
pdf_data <- with(density(sample, n = 1000), tibble(x, y))

cutoff <- 1.96

pdf_data  %>%
  ggplot(aes(x = x, y = y)) + 
  geom_line() +
  geom_ribbon(aes(ymax = y), ymin=0, fill = "red", alpha = .5, 
              data = pdf_data %>% filter(x < cutoff)) + 
  labs(x = "z score",  y = "likelihood",  title = "pdf of standard normal", subtitle = "area under curve is the probability of being less than a cutoff") + 
  annotate("text", x = 0, y = .1, label = "P(x < 1.96) = .975") + 
  scale_x_continuous(breaks = c(-1.96,  0, 1.96)) +
  theme_minimal() 
```

## What are `p` and `q`?

```{r cdf, echo = FALSE}
cdf <- ecdf(sample)

cdf_data <- 
  tibble(x = sample,
         y = cdf(sample))

cdf_data %>%
  ggplot(aes(x = x, y = y)) + 
  geom_line() +
  labs(x = "z score (aka q)", title = "cdf of standard normal (the integeral of the pdf)", y = "F(z) = Pr(x < z)  aka p") + 
  geom_hline(aes(yintercept = pnorm(cutoff)), linetype = "dashed") +
  geom_vline(aes(xintercept = cutoff), linetype = "dashed") +
  scale_x_continuous(breaks = c(-1.96,  0, 1.96)) +
  scale_y_continuous(breaks = c(0, .25, .5, .75, .975, 1)) + 
  annotate("text", x = -2, y = .9, label = "plug .975 in qnorm()\nto learn cutoff" ) +
  annotate("text", x = 1, y = .2, label = "plug 1.96 in pnorm()\nto learn P(x < q)" ) +
  theme_minimal() 
```


## What is d?

- `dnorm()`: density function, the PDF evaluated at X. 

```{r}
dnorm(0)
dnorm(1)
dnorm(-1)
```

## What is d?

`dnorm` gives the height of the distribution function. Sometimes this is called a likelihood.

```{r, echo = FALSE}


std_normal <-
  pdf_data %>%
  ggplot(aes(x = x, y = y)) + 
  geom_line() +
  theme_minimal() + 
  labs(x = "z score",  y = "likelihood",  title = "pdf of standard normal", subtitle = "d functions give height of pdf") + 
  scale_x_continuous(breaks = c(-1.96,  0, 1.96)) +
  annotate("text", x = 2.1, y = .241 + .04, label = "dnorm(1) = .24") + 
  geom_point(x = 1, y = .24, color = "blue") 

std_normal
```
