---
title: 'Lecture 9: Data Visualization (base R)'
author: "Harris Coding Camp -- Standard Track"
date: "Summer 2022"
output:
  beamer_presentation:

    includes:
      in_header: /Users/sheng-hao/Dropbox/uchicago/AIP 2nd Year/Coding Camp/Lecture
        Notes/Lecture 1/header_page_n.tex
---
```{r, setup=TRUE, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readr)
library(maps)
options(dplyr.summarise.inform = FALSE) 
housing_data <- read_csv("texas_housing_data.csv")
```

## Data Visualization: Motivation

Suppose we want to know the following info:

  - What is the annual housing sales in Texas over time?
  
  - Can we learn about the relevant housing prices of one city to another in Texas?
  
We probably want info about the *average* of housing price, how much the price *varies* over time, and info about the *extremes*

## Data Visualization: Motivation
Is simply looking at all these data helpful?

  ![](/Users/sheng-hao/Dropbox/uchicago/AIP 2nd Year/Coding Camp/Lecture
        Notes/Lecture 9/images/9_1.png)
        
## Data Visualization: Motivation
What if we make a plot of annual housing sales over time...

\medskip

```{r, echo=FALSE}
# prepare new data frame 'annual_sales':
# sum up monthly volume to get annual volume
 annual_sales <- 
  housing_data %>%
  group_by(year) %>%
  summarize(total_volume = sum(volume, na.rm = TRUE))

ggplot(data = annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_point() + geom_line()
```

## Data Visualization: Introduction

- Be able to plot in base R
- Know what types of figures you want
- Use plots to gain information

  - graphs, including axes, must be labeled properly to allow readers to understand the visualization with ease
  - variable scales should be portrayed accurately
  - graphs should be as intuitive as possible

## Data Visualization: Introduction

Two main frameworks which create graphics in R: The *base R* framework, and the *tidyverse*/*ggplot* framework

  - base R: consists of about 30 packages that are always loaded automatically when we open R
  - It is the oldest way to generate visualizations in R 
  - *base R* can and is still used to create visualizations, though most visualizations are now generated using the *tidyverse*/*ggplot* framework (more details in Lecture 10)
  
## Getting started
\footnotesize
First, let's load the data set and call it `pdat`^[For more details, see https://ladal.edu.au/introviz.html]. The data set contains:

  - `Date`: the year when a text was written
  - `Genre`: the genre of the text
  - `Text`: the name of the text
  - `Prepositions`: the relative frequency of prepositions in the text
  - `Region`: the region in which the text was written
  - `GenreRedux` collapses the existing genres into five main categories (*Conversational*, *Religious*, *Legal*, *Fiction*, and *NonFiction*) 
  - `DateRedux` collapses the dates when the texts were composed into five periods (1150-1499, 1500-1599, 1600-1699, 1700-1799, and 1800-1913)

```{r prep4}
# load online data and call it 'pdat'
pdat <- base::readRDS(url("https://slcladal.github.io/data/pvd.rda", 
                          "rb"))
```

## Getting started

Let's look at the data (first 15 rows):

  ![](/Users/sheng-hao/Dropbox/uchicago/AIP 2nd Year/Coding Camp/Lecture
        Notes/Lecture 9/images/9_2.png)

## Simplest `plot` code structure

```{r eval=FALSE}
plot(y ~ x,                  # plot y by x
     type = "[type]",        # type (points, lines, ...) 
     data = [df_name],       # data from [df_name]   
     ylab = "[y variable]",  # add y-axis label 
     xlab = "[x variable]",  # add x-axis label
     main = "[title]"        # add title 
     )                                   
```

## Understanding `plot()`: Scatter plot
\footnotesize
Let’s get started! `plot()` tells R to prepare to make a plot. 

The simplest graph is a *scatter* or *dot plot*. Scatter plots are used when the graph is set up to display the relationship between two variables.

```{r eval=FALSE}
plot(Prepositions ~ Date,   # plot Prepositions by Date
     type = "p",            # plot type "p" (points) 
     data = pdat,           # data from pdat   
     ylab = "Prepositions (Frequency)",     # add y-axis label 
     xlab = "Date (year of composition)",   # add x-axis label
     main = "Prepositions per Year"         # add title 
     )                                   
```

## Understanding `plot()`: Scatter plot
```{r eval=TRUE, echo=FALSE}
plot(Prepositions ~ Date,   # plot Prepositions by Date
     type = "p",            # plot type "p" (points) 
     data = pdat,           # data from pdat   
     ylab = "Prepositions (frequency)",     # add y-axis label 
     xlab = "Date (year of composition)",   # add x-axis label
     main = "Prepositions per Year"         # add title 
     )                                   
```

## Understanding `plot()`: Scatter plot + Vertical lines
\footnotesize
Often, we may include more info by adding some vertical lines to a plot: 

Syntax: `abline(v = [number], <other options>)`

```{r eval=FALSE}
plot(Prepositions ~ Date,   # plot Prepositions by Date
     type = "p",            # plot type "p" (points) 
     data = pdat,           # data from pdat   
     ylab = "Prepositions (frequency)",     # add y-axis label 
     xlab = "Date (year of composition)",   # add x-axis label
     main = "Prepositions per Year"         # add title 
     )     
abline(                  # add a line
  v = c(1500,1700),      # draw two v (vertical) lines
  col=c("blue", "red"),  # define line colors
  lty=c(1,2),            # define line types
  lwd=c(1,3)             # define line widths
  )   
```

## Understanding `plot()`: Scatter plot + Vertical lines

```{r eval=TRUE, echo=FALSE}
plot(Prepositions ~ Date,   # plot Prepositions by Date
     type = "p",            # plot type "p" (points) 
     data = pdat,           # data from pdat   
     ylab = "Prepositions (Frequency)",     # add y-axis label 
     xlab = "Date (year of composition)",   # add x-axis label
     main = "Prepositions per Year"         # add title 
     )     
abline(                  # add a line
  v = c(1500,1700),      # draw two vertical lines
  col=c("blue", "red"),  # define line colors
  lty=c(1,2),            # define line types
  lwd=c(1,3)             # define line widths
  )  
```

## Understanding `plot()`: Scatter plot + Horizontal lines
\footnotesize
...or by adding some horizontal lines to a plot: 

Syntax: `abline(h = [number], <other options>)`

```{r eval=FALSE}
plot(Prepositions ~ Date,   # plot Prepositions by Date
     type = "p",            # plot type "p" (points) 
     data = pdat,           # data from pdat   
     ylab = "Prepositions (frequency)",     # add y-axis label 
     xlab = "Date (year of composition)",   # add x-axis label
     main = "Prepositions per Year"         # add title 
     )     
abline(                  # add a line
  h = c(100,160),        # draw two h (horizontal) lines
  col=c("blue", "red"),  # define line colors
  lty=c(1,2),            # define line types
  lwd=c(1,3)             # define line widths
  )   
```

## Understanding `plot()`: Scatter plot + Horizontal lines

```{r eval=TRUE, echo=FALSE}
plot(Prepositions ~ Date,   # plot Prepositions by Date
     type = "p",            # plot type "p" (points) 
     data = pdat,           # data from pdat   
     ylab = "Prepositions (Frequency)",     # add y-axis label 
     xlab = "Date (year of composition)",   # add x-axis label
     main = "Prepositions per Year"         # add title 
     )     
abline(                  # add a line
  h = c(100,160),        # draw two horizontal lines
  col=c("blue", "red"),  # define line colors
  lty=c(1,2),            # define line types
  lwd=c(1,3)             # define line widths
  )   
```

## Understanding `plot()`: Scatter plot + tendency line
\footnotesize
We may consider add a linear (regression) line to the scatter plot to see the tendency, and change the color of the points:

Syntax: `abline(lm(df_name$y ~ df_name$x), <other options>)`
```{r eval=FALSE}
plot(Prepositions ~ Date,   # plot Prepositions by Date
     type = "p",            # plot type p (points) 
     data = pdat,           # data from pdat   
     ylab = "Prepositions (Frequency)",     # add y-axis label 
     xlab = "Date (year of composition)",   # add x-axis label
     main = "Prepositions per Year",        # add title 
     pch = 20,              # use point symbol 20 (filled circles)
     col = DateRedux        # define colors by DateRedux
     )
abline(                               # add a line
  lm(pdat$Prepositions ~ pdat$Date),  # draw line of linear model (lm) 
  col="red"                           # define line color as red
  )          
```

## Understanding `plot()`: Scatter plot + tendency line
```{r eval=TRUE, echo=FALSE}
plot(Prepositions ~ Date,   # plot Prepositions by Date
     type = "p",            # plot type p (points) 
     data = pdat,           # data from pdat   
     ylab = "Prepositions (Frequency)",     # add y-axis label 
     xlab = "Date (year of composition)",   # add x-axis label
     main = "Prepositions per Year",        # add title 
     pch = 20,              # use point symbol 20 (filled circles)
     col = DateRedux        # define colors by DateRedux
     )
abline(                               # add regression line (y~x) 
  lm(pdat$Prepositions ~ pdat$Date),  # draw line of linear model (lm) 
  col="red"                           # define line color as red
  )      
```

## Try it yourself
\footnotesize

1. Use the code below to load the data set and call it `dta9_1`. 

```{r ex2_1a, class.source = NULL, eval = T}
dta9_1 <-
  base::readRDS(url("https://slcladal.github.io/data/d03.rda","rb"))
```

2. Take a look at `dta9_1` -- how many variables (columns) and observations (rows) are included?

3. Find the mean and standard deviation of `Variable1` and `Variable2`.

4. Then, create a scatter plot showing `Variable1` on the x-axis and `Variable2` on the y-axis. You may change the color of the points.

5. Finally, add a linear tendency line. By eyeballing the plot, can you determine the linear relationship between the two variables?

```{r echo=FALSE, eval=FALSE}
# answer to q2
nrow(dta9_1)
ncol(dta9_1)
dta9_1

# answer to q3
mean(dta9_1$Variable1, na.rm = TRUE)
sd(dta9_1$Variable1, na.rm = TRUE)
mean(dta9_1$Variable2, na.rm = TRUE)
sd(dta9_1$Variable2, na.rm = TRUE)
  
# answer to q4 & q5
plot(Variable2 ~ Variable1, # plot Variable2 by Variable1
     type = "p",            # plot type p (points) 
     data = dta9_1,         # data from dta9_1   
     ylab = "Variable2",    # add y-axis label 
     xlab = "Variable1",    # add x-axis label
     main = "Variable2 vs Variable1",        # add title 
     pch = 20,              # use point symbol 20 (filled circles)
     col = "black"          # assign color of the points
     )
abline(                                     # add regression line (y~x) 
  lm(dta9_1$Variable2 ~ dta9_1$Variable1),  # draw line of linear model (lm) 
  col="red"                                 # define line color as red
  )  
```

## `barplot()`: Bar plot
\footnotesize
Bar plot displays frequency information across categorical variable levels. 

`barplot()` takes a table with frequency counts as its main argument. As before, we can also specify axes labels and a title. We specify text, grids, and boxes separately after `barplot()`.

```{r eval=FALSE}
barplot(table(pdat$DateRedux),        # plot Frequency by DateRedux
     ylab = "Frequency",              # add y-axis label 
     xlab = "Period of composition",  # add x-axis label 
     main = "bar plot in base R",     # add title
     col = c("grey", "pink", "green", "lightgrey", "steelblue"),
                                      # define color of each bar
     ylim = c(0, 250)                 # define y-axis limits
     )                                    
text(seq(0.7, 5.5, 1.2),              # add label positions (x-axis)
     table(pdat$DateRedux)+10,        # add label positions (y-axis)
     table(pdat$DateRedux))           # add labels
box()                                 # add box
```

## `barplot()`: Bar plot

```{r eval=TRUE, echo=FALSE}
barplot(table(pdat$DateRedux),        # plot Frequency by DateRedux
     ylab = "Frequency",              # add y-axis label 
     xlab = "Period of composition",  # add x-axis label 
     main = "bar plot in base R",     # add title
     col = c("grey", "pink", "green", "lightgrey", "steelblue"),
                                      # define color of each bar
     ylim = c(0, 250)                 # define y-axis limits
     )                                    
text(seq(0.7, 5.5, 1.2),              # add label positions (x-axis)
     table(pdat$DateRedux)+10,        # add label positions (y-axis)
     table(pdat$DateRedux))           # add labels
box()                                 # add box
```

## `barplot()`: Grouped bar plot
\footnotesize
To create grouped bar plots, we tabulate the variables that we are interested in (e.g. Region)
```{r eval=FALSE}
barplot(table(pdat$DateRedux, pdat$Region),  # plot Frequency 
                                             # by DateRedux and Region
        beside = T,                          # bars beside each other
        ylab = "Frequency",                  # add y-axis label 
        xlab = "Period of composition",      # add x-axis label
        main = "grouped bar plot in base R", # add title
        col = c("grey", "pink", "green", "lightgrey", "steelblue"),
                                             # define color of each bar
        ylim = c(0, 250)                     # define y-axis limits
        )                                    
text(c(seq(1.5, 5.5, 1.0), seq(7.5, 11.5, 1.0)), # add label positions (x-axis)
     table(pdat$DateRedux, pdat$Region)+10,      # add label positions (y-axis)
     table(pdat$DateRedux, pdat$Region))         # add labels
legend("topleft", names(table(pdat$DateRedux)))  # add legend                    
box()                                            # add box
```

## `barplot()`: Grouped bar plot
```{r eval=TRUE, echo=FALSE}
barplot(table(pdat$DateRedux, pdat$Region),  # plot Frequency by DateRedux and Region
        beside = T,                          # bars beside each other
        ylab = "Frequency",                  # add y-axis label 
        xlab = "Period of composition",      # add x-axis label
        main = "grouped bar plot in base R", # add title
        col = c("grey", "pink", "green", "lightgrey", "steelblue"),
                                             # define color of each bar
        ylim = c(0, 250)                     # define y-axis limits
        )                                    
text(c(seq(1.5, 5.5, 1.0), seq(7.5, 11.5, 1.0)),    # add label positions (x-axis)
     table(pdat$DateRedux, pdat$Region)+10, # add label positions (y-axis)
     table(pdat$DateRedux, pdat$Region))    # add labels
legend("topleft", names(table(pdat$DateRedux))) # add legend                    
box()                                       # add box
```

## `boxplot()`: box plot
\footnotesize
Boxplots show the relationships between categorical and numeric variables. 

They are very useful because they not only provide measures of central tendency (e.g. median) but also offer info about the distribution (e.g. 1st and 3rd quartile, min, max) of the data.

  ![](/Users/sheng-hao/Dropbox/uchicago/AIP 2nd Year/Coding Camp/Lecture
        Notes/Lecture 9/images/9_3.png)

## `boxplot()`: box plot
```{r eval=FALSE}
boxplot(Prepositions ~ DateRedux, data = pdat)                       
```

```{r eval=TRUE, echo=FALSE}
boxplot(Prepositions ~ DateRedux, data = pdat)                       
```

## `boxplot()`: box plot
\footnotesize
We can change the color of the borders and/or fill colors in each box:
```{r eval=FALSE}
boxplot(Prepositions ~ DateRedux, data = pdat,
        border = c("blue", "red", "steelblue", "black", "green"),
        col = c("grey", "pink", "green", "lightgrey", "steelblue"))                         
```

## `boxplot()`: box plot
```{r eval=TRUE, echo=FALSE}
boxplot(Prepositions ~ DateRedux, data = pdat,
        border = c("blue", "red", "steelblue", "black", "green"),
        col = c("grey", "pink", "green", "lightgrey", "steelblue"))                           
```

## Try it yourself: Plot 1
\footnotesize
1. First, load the dataset `weight_chart.txt` and give it a name (e.g. weight_chart). Don't forget to change the working directory if needed! Then, take a look at the dataset.
```{r eval=FALSE}
# read the space separated dataframe
weight_chart <- read.delim("weight_chart.txt") 
weight_chart
```
2. Then, try to replicate the plot in the next slide by modifying the following code. Hint: When choosing type, use `"b"` (which means *both* line and points).
```{r eval=FALSE}
plot(??? ~ ???,             # fill in appropriate X and Y
     type = "b",            # plot type "b" (both line and points) 
     pch = 15,              # filled square
     cex = 1.5,             # size of pch symbols
     data = ???,            # data source
     ylim = c(?,?),         # fill in the y-axis limits 
     ylab = "???",    
     xlab = "???",    
     main = "???"
     )                           
```

## Try it yourself: Plot 1
```{r results="hide", eval=TRUE, echo=FALSE}
# read the space separated dataframe
weight_chart <- read.delim("weight_chart.txt") 
# view data frame
weight_chart
```

```{r eval=TRUE, echo=FALSE}
# answer
plot(
  Weight ~ Age,
  type="b",
  pch=15,
  cex=1.5,
  data = weight_chart,
  ylim=c(2,10),
  xlab="Age (months)",
  ylab="Weight (kg)",
  main="Weigh gain during early infant development"
)
```

## Try it yourself: Plot 2
\footnotesize
1. First, load the dataset and give it a name (e.g. `pdat`). 
```{r eval=FALSE}
# load data and call it 'pdat'
pdat <- base::readRDS(url("https://slcladal.github.io/data/pvd.rda", 
                          "rb"))
```
2. Then, try to replicate the plot in the next slide by modifying the following code. 
```{r eval=FALSE}
barplot(table("_fill in_"),       
     ylab = "[_fill in_]",              
     xlab = "[_fill in_]",     
     main = "[_fill in_]",       
     ylim = c(?, ?),     # fill in the y-axis limits
     col = rainbow(5)    # a built-in color palettes which can be used 
                         # to quickly generate color vectors
     )                                    
box()                      
```

## Try it yourself: Plot 2
```{r eval=TRUE, echo=FALSE}
# answer
barplot(table(pdat$GenreRedux),       
     ylab = "Frequency",              
     xlab = "Main Genres",     
     main = "bar plot: GenreRedux",       
     ylim = c(0, 250),
     col = rainbow(5)
     )                                    
box()                             
```

## Try it yourself: Plot 3
\footnotesize
1. Use the dataset `pdat` and try to replicate the plot in the next slide by modifying the following code. For the color of each box, you can choose whichever you want.
```{r eval=FALSE}
boxplot(Prepositions ~ ???, data = ???,
        col = ???                         
```

## Try it yourself: Plot 3
```{r eval=TRUE, echo=FALSE}
# answer
boxplot(Prepositions ~ GenreRedux, data = pdat,
        col = c("grey", "pink", "green", "purple", "blue"))                        
```

## Appendix: Histogram
\footnotesize
It summarizes numeric variables by showing their distribution across bins.

```{r eval=TRUE}
hist(pdat$Prepositions,
     xlab = "Prepositions (per 1,000 words)", main = "")
```

## Appendix: Density Plot
\footnotesize
It is a smoothed version of the histogram and is used in the same concept.

```{r eval=TRUE}
plot(density(pdat$Prepositions),
     xlab = "Prepositions (per 1,000 words)", main = "")    
```

## Appendix: Line Plot
\footnotesize
It is particularly useful in time series or finance.

```{r eval=TRUE}
plot(pdat$Prepositions,
     type = "l",
     ylab = "Prepositions", main = "")    
```

## Recap

- Visualizing our data can help lead to powerful insights between variable relationships

  - Making quick plots helps us understand data and makes us aware of data issues

- We've learned how to make commonly used plots, including:

  - scatter plots (with auxiliary lines)
  - bar plots
  - box plots

- There are many ways you can visualize your data

  - We can even use `ggplot()` to generate similar plots more conveniently! (more details in Lecture 10) 